Toward dynamic models of combat or lack thereof in animal mating

### 2016-01-18 ms.tex
added naive event algorithm to ms

\subsection{Algorithms}
Simulating the population dynamics requires accounting for four types of events:
\begin{itemize}
    \item Immature individuals become mature
    \item Searching males encounter an empty nest
    \item Searching males encounter an occupied nest and contest it
    \item Occupying males abandon an occupied nest
\end{itemize}

@@

\subsection{Event Algorithm}
See Fig. \ref{fig:events}. Population state generates events, events are either:
\begin{enumerate}
    \item A new individual x has matured and entered the pool (based off of a trait(?))
    \item Discovery event wherin both a nest RR (repoductive resource) and an individual x are drawn from a pool, the metabolic costs of searching are deducted from the time differences between events, that nest is either:
    \begin{enumerate}
        \item un-occupied: in which x takes ownership of RR.
        \item occupied by x' in which a constest occurs (see fig 2)
    \end{enumerate}
\end{enumerate}

@@

these species have $R_\mu$ different interactions
\begin{table}[h!]
    \caption{state changes}
    \centering
    \begin{tabular}{c | c | c | c | l }
        \hline
        Reaction num & reactants  & & products & Description\\
        \hline
        \hline
            $R_1$       &   $S_1$       & $--c_1-->$      &   $ S_2$      & maturation\\
            $R_2$       &   $S_2 + S_3$ & $--c_2-->$      &   $S_4 $      & occupation\\
            $R_3$       &   $S_2 + S_4$ & $--c_3-->$      &   $S_4 + S_5$ & contest\\
            $R_4$       &   $S_5$       & $--c_4-->$      &   $S_2$       & recovery\\
            $R_5$       &   $S_2 + S_4$ & $--c_5-->$      &   $S_4$       & contest with death\\
            $R_6$       &   $S_4$       & $--c_6-->$      &   $S_2 + S_3$ & abandon event\\
            $R_7$       &   $S_1$       & $--c_7-->$      &   $0$         & metab0lic / predation event\\
            $R_8$       &   $S_2$       & $--c_8-->$      &   $0$         & predation event\\
            $R_9$       &   $S_4$       & $--c_9-->$      &   $S_3$       & metab0lic / predation event\\
            $R_{10}$    &   $S_6$       & $--c_{10}-->$   &   $S_4$       & nest handover\\
        \hline
    \end{tabular}
\end{table}

$c_n$ is the stochatic reaction constant\\
$c_n \cdot dt = $ the average probability that a particular combination of $R_n$, reactant molecules will react accordingly in the next infinitesimal time interval $dt$.
\citep{Gillespie-1976}

we generate the reactions then step through them with a population.

\subsection{Contest Algorithm}
See Fig. \ref{fig:contests}.
For a defender $x_{def}$ vs an attacker $x_{atk}$, there are 3 stages:
each individual has an aggression trait $a$ and a mass $m$.
\begin{equation}\label{eq:logit}
    logit^{-1}(\alpha) = \frac{exp(\alpha)}{1 + exp(\alpha)}
\end{equation}

$m$ is mass
$a$ is aggression trait
\begin{enumerate}
    \item Display 1
        Outcomes:
        \begin{itemize}
            \item $x_{atk}$ escalates: 
                $Pr(\rm{escalation}) =  logit^{-1}(a_{atk} \cdot m_{atk} - m_{def})$
                or leaves.
            \item $x_{def}$ escalates:
                $Pr(\rm{escalation}) =  logit^{-1}(a_{def} \cdot m_{def} - m_{atk})$ 
                or leaves.
            \item if the attacker leaves the defender keeps the nest
            \item if the defender leaves the attacker keeps the nest
            \item the attacker is evaluated first
            \item if they both escalate then the next stage occurs
        \end{itemize}
    \item Display 2:
        similar to stage 1, this could involve the $R$ value of the nest, or include the oponents aggression levels or energy reseves(?).
    \item Fight
        So far I have something like
        \begin{itemize}
            \item $x_{def}$ wins:
            $Pr(\rm{def}) = logit{-1}(m_{def} - m_{atk})$
        \end{itemize}
\end{enumerate}
The cost of the fighting stage is much greater than the cost of the display stage.

>>>>

\subsection{Algorithms}
\subsection{Event Algorithms}
So far we are considering two algorithms for stepping through events one is a stochastic simulation algorithm and a simpler (although more compuationally expensive) stepping algorithm.
\subsection{SSA}
Simulating the population dynamics requires accounting for four types of events:
\begin{itemize}
    \item Immature individuals become mature
    \item Searching males encounter an empty nest
    \item Searching males encounter an occupied nest and contest it
    \item Occupying males abandon an occupied nest
\end{itemize}

@@

See Fig. \ref{fig:events}. Population state generates events, events are either:
\begin{enumerate}
    \item A new individual x has matured and entered the pool (based off of a trait(?))
    \item Discovery event wherin both a nest RR (repoductive resource) and an individual x are drawn from a pool, the metabolic costs of searching are deducted from the time differences between events, that nest is either:
    \begin{enumerate}
        \item un-occupied: in which x takes ownership of RR.
        \item occupied by x' in which a constest occurs (see fig 2)
    \end{enumerate}
\end{enumerate}

@@

these species have $R_\mu$ different interactions
\begin{table}[h!]
    \caption{state changes}
    \centering
    \begin{tabular}{c | c | c | c | l }
        \hline
        Reaction num & reactants  & & products & Description\\
        \hline
        \hline
            $R_1$       &   $S_1$       & $--c_1-->$      &   $ S_2$      & maturation\\
            $R_2$       &   $S_2 + S_3$ & $--c_2-->$      &   $S_4 $      & occupation\\
            $R_3$       &   $S_2 + S_4$ & $--c_3-->$      &   $S_4 + S_5$ & contest\\
            $R_4$       &   $S_5$       & $--c_4-->$      &   $S_2$       & recovery\\
            $R_5$       &   $S_2 + S_4$ & $--c_5-->$      &   $S_4$       & contest with death\\
            $R_6$       &   $S_4$       & $--c_6-->$      &   $S_2 + S_3$ & abandon event\\
            $R_7$       &   $S_1$       & $--c_7-->$      &   $0$         & metab0lic / predation event\\
            $R_8$       &   $S_2$       & $--c_8-->$      &   $0$         & predation event\\
            $R_9$       &   $S_4$       & $--c_9-->$      &   $S_3$       & metab0lic / predation event\\
            $R_{10}$    &   $S_6$       & $--c_{10}-->$   &   $S_4$       & nest handover\\
        \hline
    \end{tabular}

\end{table}

$c_n$ is called the stochatic reaction constant\\
$c_n \cdot dt = $ the average probability that a particular combination of $R_n$, reactant molecules will react accordingly in the next infinitesimal time interval $dt$

\citep{Gillespie-1976}

We generate the reaction events then step through them with a population.
\clearpage
\subsection{Naive Event Algorithm}
we have 3 lists
\begin{verbatim}
    immature males
    searching males
    nests
\end{verbatim}

every full time unit that the searching male is alive for it has a chance to discover a nest based off of its exploration trait.

\begin{verbatim}
time = 0
while time < time when the females mature:
    check which males have matured
        add them to the list of searching males
    for each occupied nest:
        deduct metabolic costs from the occupier
        remove dead males

    for each searching male m:
        deduct metabolic costs from m
        remove dead males
        reduce time until next event for m
        if time until next event <= 0:
            reset the time until next event
            if random > exploration_prob:
                choose a nest at random
                if it is occupied:
                    contest the occupying male for control of the nest
                    if m has won the contest:
                        m occupy's n
                        the loser is added to the list of searching males
                else:
                    occupy the nest
    t += delta_time
\end{verbatim}

\subsection{Contest Algorithm}
See Fig. \ref{fig:contests}.
For a defender $x_{def}$ vs an attacker $x_{atk}$, there are 3 (currently 2) stages:
each individual has an aggression trait $a$ and a mass $m$.
\begin{equation}\label{eq:logit}
    logit^{-1}(\alpha) = \frac{exp(\alpha)}{1 + exp(\alpha)}
\end{equation}

$m$ is mass
$a$ is aggression trait
\begin{enumerate}
    \item Display 1
        Outcomes:
        \begin{itemize}
            \item $x_{atk}$ escalates: 
                $Pr(\rm{escalation}) =  logit^{-1}(a_{atk} \cdot m_{atk} - m_{def})$
                or leaves.
            \item $x_{def}$ escalates:
                $Pr(\rm{escalation}) =  logit^{-1}(a_{def} \cdot m_{def} - m_{atk})$ 
                or leaves.
            \item if the attacker leaves the defender keeps the nest
            \item if the defender leaves the attacker keeps the nest
            \item the attacker is evaluated first
            \item if they both escalate then the next stage occurs
        \end{itemize}
    \item Display 2:
        similar to stage 1, this could involve the $R$ value of the nest, or include the oponents aggression levels or energy reseves(?).
    \item Fight
        So far I have something like
        \begin{itemize}
            \item $x_{def}$ wins:
            $Pr(\rm{def}) = logit^{-1}(m_{def} - m_{atk})$
        \end{itemize}
\end{enumerate}
The cost of the fighting stage is much greater than the cost of the display stage.
